Cultural Transmission and Technological Transitions during the Late Paleolithic Period in Korea

### 2023-02-17 analysis/paper/paper.qmd
Edited based on proof-reading

---
title: "Cultural Transmission and Technological Transitions during the Late Paleolithic in Korea"
author:
  - Gayoung Park:
      institute: uw
      correspondence: "yes"
      email: gayoungp@uw.edu
  - Ben Marwick:
      institute: uw
institute:
  - uw:
      name: Department of Anthropology, University of Washington
      address: Seattle, WA, USA
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
  cache: false
editor: visual
date: now
date-format: long
title-block-published: "Last updated" 
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style

abstract: |
  The onset of the Late Paleolithic period in Korea, represented by the appearance of stemmed points and blades, was a key event in the dispersal of modern humans in Northeast Asia. Previous studies have mainly focussed on possible origin locations of these new technologies. The specific cultural processes of the appearance stemmed points and blades has rarely been considered. We investigate the cultural processes by applying a cultural transmission framework to investigate the social contexts of the emergence of these new technologies. Our main question is: what was the dominant mode of cultural transmission during this time of technological innovation in the Korean Late Paleolithic? Following Bettinger and Eerkens [-@bettinger1999point], we evaluated models of guided variation and indirect bias using data from Korean assemblages containing stemmed points. To evaluate these models and understand the transmission processes, we computed correlation coefficients and coefficients of variation (CV). We found that information about the new technology was likely transmitted via guided variation with small impact of indirect bias. Some attributes including length and width were transmitted with less variation while other attributes appear to have more variation. Our results suggest that the dominant mode of cultural transmission for the earliest stemmed points was guided variation. A slight shift in the social context of transmission towards indirect bias was also observed in the later chronological phase. We assume that individuals or groups developed stemmed points by experimenting with existing blade technologies and then copied crucial parts of a successful model to ensure the quality to optimize tool usage. As a result, the shape of stemmed points became more standardized among their social groups.

@@

The application of evolutionary theory to archaeological research has been favored for the study of technological transitions and related human behaviors in the remote past [@dunnell1980evolutionary; @lipo1997population; @bettinger1997rediscovering; @bettinger1999point; @mesoudi2008cultural].  Archaeologists have used evolutionary theories and methods to study human behavioral ecology, cultural transmission, and artifact phylogenetics in the past [@riede2010isn; @o2017dual; @garvey2018current]. In this paper we use cultural transmission theory to investigate technological transitions during the Korean Late Paleolithic. The primary technological innovation of this period was the introduction of stemmed points and blades. These new lithic technologies may represent the first arrival of modern humans in this region, and thus they may also mark an important event in human dispersal through East Asia [@seong2009emergence]. Most previous studies on stemmed points focused on where they may have originated, connecting Korea with global patterns of modern human dispersal. There are, however, largely unanswered questions about the specific cultural processes and social contexts of this technological change in Korea [@seong2008tanged; @bae2010origin; @bae2017origin]. Our study explores the social contexts in which new technologies emerged in the Korean Late Paleolithic based on a cultural transmission framework. Our main question is: What was the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic? We also ask: Do modes of cultural transmission vary over time and space? We consider three possible modes of cultural transmission: guided variation (trial and error), indirect bias (copying others), or a combination of the two. The results of this study have implications for determining whether these novel technologies originated outside of the Korean Peninsula or if they were locally developed independently.

@@

The emergence of stemmed points marks the beginning of the Late Paleolithic period in Korea around 40-35 ka. A stemmed point is a projectile point made out of an elongated flake or blade with a slight retouch on the proximal end to shape an acute tip, and on the distal end to make a stem, which connects to a shaft. Stemmed points were the first composite tool types to appear on the Korean Peninsula, and were a symbol of new hunting strategies there and in adjacent regions [@seong2008tanged; @lee2019were]. A number of technological innovations are evident on stemmed points, including blade technology, multiple manufacturing stages, and evidence of resharpening and reusing, which were rarely seen earlier [@chang2013human; @bamforth2009projectile; @seong2015diversity]. Currently, the oldest stemmed points in Northeast Asia are from the Yonghodoing site in Korea, dating back to 38.5ka [@seong2008tanged; @seong2015diversity; @bae2012nature]. Following their appearance in Korea, stemmed points spread to the Japanese archipelago [@chang2013human].

Previous studies of the Late Paleolithic technological transitions in Korea have mainly focused on the possible origin locations of stemmed points, as part of the discussion of the timing and routes of modern human dispersal in eastern Asia. The debate about the origin of stemmed points can be summarized into two competing models: *in situ* evolution [@seong2009emergence] and heterogenic migration [@bae2010origin]. The *in situ* model claims that stemmed points and other Late Paleolithic assemblages including blade industries autonomously emerged in the Korean peninsula, as a form of convergent evolution [@seong2006structure; @seong2008tanged; @seong2009emergence]. To support his claim, @seong2009emergence examined the blade-to-flake ratios of stone artifact assemblages in South Korea. In his view, the blade industry represents a new technology while flakes indicate a continuously-used existing one. He argues that the increased ratio of blades in stone artifact assemblages during the Late Paleolithic shows an expansion of the new technology after its local invention. In addition, Seong claims that increased numbers of stemmed points over time, and standardization of their shape, supports the prominance of gradual, local, evolutionary processes in the emergence of new technologies.

In contrast, the migration model argues that the new blade industry, including stemmed points, and the earlier simple flake tool tradition, including large cores, polyhedrals, choppers and handaxes, came from different origins outside of the Korean peninsula [@bae2010origin; @bae2012nature; @bae2013early]. While the *in situ* model claims that the heterogenic character is the result of indigenous development, the migration model proposes that it is the result of the continuous influx of modern human populations from both north and south. Specifically, the blade technology is claimed to have been introduced from Siberia, Mongolia, or other regions of northeast China following the Liaohe and Sunghe rivers around 35 ka BP, while the simple flake-tool tradition came from southern China [@bae2013early]. The migration of people from the southern route is supported by genetic studies of the Y chromosome, indicating that the O3-M122 M122 haplotype originated from southern East Asia and moved to northeastern Asia, including Korea, at 30-25 ka BP [@shi2005chromosome]. Bae et al. [-@bae2012nature] assume this southern migration could be related to paleoenvironmental fluctuations during the MIS 3 to 2 transition, which made the Yellow Sea/West Sea region open.

Lee [-@lee2013current] argues that the transition to the Korean Late Paleolithic might be more complicated than either models of migration or *in situ* development. He partly agrees with the *in situ* model that simple flake tools had continuously been used in Korea as the result of ancestor-descendant relationships, under conditions of low effective population size. With regard to the blade industry, he claims that low degrees of uniformity and small quantities of blade-associated toolkits indicate an origin outside of Korea, perhaps resulting from trade or migration. We explore these three options, *in situ*, migration, and a mixture of the two, by measuring transmission biases in assemblages of stemmed points.

@@

To measure transmission biases in tool-making, we draw on Darwinian evolutionary theory. This body of theory has helped archaeologists understand a variety of technological innovations and related human behaviors, for example using human behavioral ecology, phylogenetics, cultural transmission theory, and niche construction theory [@dunnell1980evolutionary; @richerson1992cultural; @lipo1997population; @bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural]. Cultural evolutionary approaches are effective at explaining cultural changes using mechanisms of inheritance, variation-generating processes, and selection. The mechanisms are similar to biological evolution, but in cultural systems they have unique and distinctive properties. Social transmission strongly affects these key evolutionary processes [@Whiten_2017], and can occur through various learning processes, such as stimulus enhancement, emulation, imitation, and teaching [@o2000applying; @schillinger2014copying]. We focus here on how different social contexts can result in different modes of social transmission of tool-making skills among hunter-gather populations.

Cultural transmission theory holds that information about social and technological behaviours is acquired through interaction with other individuals and the environment [@boyd1988culture; @richerson1992cultural]. Individuals learn by themselves (e.g. trial and error), or from each other by sharing information. Information can be modified (also known as 'biased') depending on an individual's transmission context and cultural repertoire. Modification of information can occur by recombination, loss, or partial alteration [@eerkens2005cultural; @o2017dual]. Transmission biases can be important loci of changes in material culture, and can be influenced by the social contexts of cultural transmission [@heyes1994social; @eerkens2007cultural; @lycett2015cultural; @creanza2017cultural; @kendal2018social]. These biases include guided variation (where individuals learn new behaviors through trial and error), content-based bias (where some aspect of the transmitted instructions, such as cultural preferences, makes them more likely to be adopted), frequency-based bias (where an individual is biased to choose particular behaviors based on their perceived frequency in the population, such as extremely popular or rare behaviors), and indirect bias (where a behavior is transmitted because of its association with other attributes, such as the prestige or skill of other individuals) [@boyd1988culture; @richerson1992cultural; @o2017dual].

@@

Following Bettinger and Eerkens' framework, Garvey [-@garvey2018current] uses simulation to explore the degree of standardization represented by coefficient of variation (CV) values of projectile points from the US Southwest and westernmost southern High Plains. Garvey measured CV values for weight, thickness, width and length on two types of projectile points, Washita and Fresno points, from the Henderson site. She then computed simulated CV values according to three scenarios of different levels of transmission fidelity (i.e. CV = 10%, 5%, and 3%). The observed metrics of archaeological projectile points are closest to simulated metrics with 3% CV, which represents "extremely high-fidelity copying". A CV of 3% is the Weber fraction, or threshold of human visual perception, with variation below this value being too subtle for people to notice [@eerkens2000practice]. Garvey's work demonstrates how the cultural transmission of tool-making behaviors can be measured, simulated, and interpreted from the archaeological record.

Garvey and Bettinger and Eerkens inferred transmission processes from material culture because they were unable to observe them directly. Direct observations have been reported by Mesoudi and O'Brien [-@mesoudi2008cultural] who conducted experimental research using groups of undergraduate students to test the underlying assumptions of Bettinger and Eerkens' study. They simulated model-based bias in projectile point making by providing the design of a model and information about the model's prior success to their research participants. They also simulated guided variation by allowing their participants to explore their own designs. They observed that the majority of participants who were able to choose the previous design copied the most-successful model. Metric attributes of projectile points made when copied from successful models were more highly correlated than attributes of points made by trial and error. Mesoudi and O'Brien's experimental results are critical because they confirm the robustness of previous assumptions about cultural transmission biases, and validate the connection between material culture variation and cultural transmission biases.

@@

Inspired by Bettinger and Eerkens' approach, we use two contrasting transmission modes, guided variation and indirect bias, to investigate the spread of stemmed point technology during the Korean Late Paleolithic period. We propose a spectrum on which we can locate foraging groups and how they started to make stemmed points, depending on their degree of social isolation or social connectedness. On one end of the spectrum we have socially isolated groups who made stemmed points through guided variation; and on the other end of the spectrum we have socially connected groups whose knowledge of stemmed points derived from transmission processes dominated by indirect bias.

To contextualize this further, on one end of our spectrum we have socially isolated groups who stayed in physically remote places from other groups, or had unfamiliar neighbors with limited contact between groups. Our assumption for this social context is that individuals or groups acquired the technology of stemmed points by modifying existing flake tool forms through trial-and-error processes to solve problems relating to resource procurement (cf. Seong's *in situ* model). If this social context was prevalent, we predict that this trial and error behavior could have left a distinctive signature on the metric variables of the stemmed points. If trial and error was the dominant bias in the transmission of knowledge about how to make stemmed points, we expect that the morphological attributes on a stemmed point will be poorly correlated and show high variation within assemblages and between sites.

On the other end of our spectrum, we have socially connected groups that occupied places close to other groups or had regular contact with them. Relatively high degrees of social connectivity provide frequent opportunities for observing others and acquiring information. Individuals or groups learned the technology of stemmed points by copying a model from another individual or group. In this scenario, it is likely that the model, or ideal, stemmed point has been highly successful and frequently chosen, and learners copy all information about the point design as a package. This implies a small number of locations, or a single point (perhaps from a source external to the Korean peninsula, cf. Bae's migration model), where stemmed points first appeared and then spread from. In this social context, indirect bias is the dominant influence on the transmission of lithic technology. If indirect bias was prominent in the Korean Paleolithic, we expect that stemmed points would be more standardized, attributes on the stemmed points would be more correlated, and assemblages from multiple sites would exhibit less variance between and within them.

@@

After the first discovery of stemmed points at the Seokjangri site in the 1960s, more than 450 have been found in nearly 30 sites across Korea (@fig-map) [@Sohn_1967; @lee2019were; @Chong_2021]. While most sites contain only a few points, a few sites have many more, such as Suyanggae (n = 55), Jingeuneul (n = 99), and Yongsandong (n = 38) [@Kim_2017]. Among these stemmed points, we selected those that are unbroken from the tip to the stem. We excluded artifacts that were recorded as stemmed points but lack a stem. This resulted in a sample of `r total_number_of_sp` stemmed points from `r total_number_of_assemblages` assemblages unearthed from `r total_number_of_sites` sites spanning the period 40-17 ka (@tbl-sp-number). The images of the stemmed points were obtained through published excavation reports and direct photography during our research in local museums. We defined multiple assemblages from a site where artifact-bearing deposits were separated by culturally sterile deposits, or where distinct artifact-bearing stratigraphic units could be identified by major differences in the texture, color, and composition of the sedimentary deposits.

@@

To analyze morphological change over time, we used previously developed chronologies that identify three phases in the Korean Late Paleolithic [@seong2015diversity; @Park_2013]. These chronologies were constructed based on radiocarbon ages, and assemblages without radiocarbon ages were classified by blades, stemmed point blanks, and toolkit composition. Following these previous chronological schemes, we divided the Korean Late Paleolithic assemblages into three chronological phases: 1) stemmed points made out of flakes and no blades in assemblages, 2) stemmed points made out of blades or flakes and the existence of blades in assemblages, and 3) stemmed points made out of blades and the existence of micro blades in assemblages (@tbl-korean-chronology). Applying this division is necessary because some sites such as Yonghodong, Goryeri, Jungmal, and Mungyeong have no radiocarbon ages due to research limitations. Inferring ages for these sites by analogy to the technological sequences at sites with dates is necessary for making maximal use of the available archaeological data to increase our sample size. We arranged our `r total_number_of_assemblages` assemblages containing stemmed points into these three phases to facilitate observations of change over time.

@@

Panel A of @fig-raw-materials-by-phase summarizes the amount of stemmed points in each chronological phase, and the distribution of raw materials across the phases. We exclude phase 1 from our analyses below because of the small sample size to compute correlations and CVs. Shale is the dominant raw material in both Phase 2 and 3. In Phase 3 there is an increase in the proportion of porphyry, driven largely by finds from Yongsangdong, and other raw materials. Panel B of @fig-raw-materials-by-phase shows the count of stemmed points by raw material type in assemblages with more than five stemmed points. Much of the raw material diversity comes from isolated finds, with those assemblages with more than five points showing high homogeneity in raw materials.

```{r}
#| label: fig-raw-materials-by-phase
#| fig-cap: "A: Three chronological phases of the Korean Late Paleolithic, and frequecies of stemmed points and distribution of raw materials in each phase. B: Frequencies of stemmed points by assemblage and raw material, for assemblages with five or more stemmed points."
#| fig-height: 4

@@

Following the metric attributes used in previous studies of cultural transmission and projectile points, i.e. maximum length and width [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we examined variations in shape and the relationship between each attribute using morphological attributes on stemmed points (@fig-labelled-schematic). We obtained our morphological attribute data from landmark analysis of digitized images of stemmed points. Compared to traditional caliper measurements, morphometric data yield more easily interpretable numerical and visual outcomes [@buchanan2010geometric; @thulman2012discriminating; @cardillo2016environment; @petvrik2018shape; @macleod2018quantitative; @okumura2019archaeology; @suarez2019life]. We did not include weight and thickness since these measurements were not available to us. For the landmark analysis, we placed 11 landmarks on the outline of each stemmed point and calculated distances between landmark coordinate pairs to derive attributes for statistical analysis. The landmarks we recorded are described in @tbl-landmark-abbreviations and shown in @fig-labelled-schematic. Using the point tool in ImageJ [@schneider2012nih] we captured the landmarks from images of the artifacts, and exported them as XY coordinate data for further analyses.

@@

We chose Coefficients of Variation (CV) to measure the variation among the measurements of the stemmed points. The CV is determined by the ratio of the standard deviation to the mean (usually expressed as a percentage). The method has been used in various disciplines to calculate standardization, precision, equality, homogeneity, etc. [@ng2006performance; @panichkitkosolkul2009improved; @panichkitkosolkul2013confidence; @wang2020standardization]. In archaeological studies, CV has been applied to measure the variation between artifacts and to test hypotheses about cultural evolutionary processes, including distinguishing between the types of learning biases affecting cultural transmission [@eerkens2001techniques; @eerkens2005cultural; @schillinger2014copying; @garvey2018current]. As a guide to interpreting CV values, Eerkens and Bettinger [-@eerkens2001techniques] claim that over 57.7% of the CV is the result of random production and below 1.7% is the by-product of using a scale or template. Among all the various artifacts for which \@eerkens2001techniques summarized CV values, the most comparable to stemmed points are Great Basin projectile points, whose CV values range from 6-55%, with an average of 22%. Similarly, projectile points from the US Southwest have CV values ranging from 11% to 33% [@garvey2018current].

>>>>

---
title: "Cultural Transmission and Technological Transitions during the Late Paleolithic Period in Korea"
author:
  - Gayoung Park:
      institute: uw
      correspondence: "yes"
      email: gayoungp@uw.edu
  - Ben Marwick:
      institute: uw
institute:
  - uw:
      name: Department of Anthropology, University of Washington
      address: Seattle, WA, USA
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
  cache: false
editor: visual
date: now
date-format: long
title-block-published: "Last updated" 
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style

abstract: |
  The onset of the Late Paleolithic period in Korea, represented by the appearance of stemmed points and blades, was a key event in the dispersal of modern humans in Northeast Asia. Previous studies have mainly focused on possible origin locations of these new technologies. The specific cultural processes of the appearance of stemmed points and blades have rarely been considered. We investigate these cultural processes by applying a cultural transmission framework to investigate the social contexts of the emergence of these new technologies. Our main question is: what was the dominant mode of cultural transmission during this time of technological innovation in the Korean Late Paleolithic? Following Bettinger and Eerkens [-@bettinger1999point], we evaluated models of guided variation and indirect bias using data from Korean assemblages containing stemmed points. To evaluate these models and understand the transmission processes, we computed correlation coefficients and coefficients of variation (CV). We found that information about the new technology was likely transmitted via guided variation with small impact of indirect bias. Some attributes, including length and width, were transmitted with less variation while other attributes appear to have more variation. Our results suggest that the dominant mode of cultural transmission for the earliest stemmed points was guided variation. A slight shift in the social context of transmission towards indirect bias was also observed in the later chronological phase. We assume that individuals or groups developed stemmed points by experimenting with existing blade technologies and then copied crucial parts of a successful model to ensure the quality to optimize tool usage. As a result, the shape of stemmed points became more standardized among their social groups.

@@

The application of evolutionary theory to archaeological research has been favored for the study of technological transitions and related human behaviors in the remote past [@dunnell1980evolutionary; @lipo1997population; @bettinger1997rediscovering; @bettinger1999point; @mesoudi2008cultural].  Archaeologists have used evolutionary theories and methods to study human behavioral ecology, cultural transmission, and artifact phylogenetics in the past [@riede2010isn; @o2017dual; @garvey2018current]. In this paper, we use cultural transmission theory to investigate technological transitions during the Korean Late Paleolithic. The primary technological innovation of this period was the introduction of stemmed points and blades. These new lithic technologies may represent the first arrival of modern humans in this region, and thus, they may also mark an important event in human dispersal through East Asia [@seong2009emergence]. Most previous studies on stemmed points focused on where these new technologies may have originated, connecting Korea with global patterns of modern human dispersal. There are, however, largely unanswered questions about the specific cultural processes and social contexts of this technological change in Korea [@seong2008tanged; @bae2010origin; @bae2017origin]. Our study explores the social contexts in which new technologies emerged in the Korean Late Paleolithic based on a cultural transmission framework. Our main question is: What was the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic? We also ask: Do modes of cultural transmission vary over time and space? We consider three possible modes of cultural transmission: guided variation (trial and error), indirect bias (copying others), or a combination of the two. The results of this study have implications for determining whether these novel technologies originated outside of the Korean Peninsula or if they were locally developed independently.

@@

The emergence of stemmed points marks the beginning of the Late Paleolithic period in Korea around 40-35 ka. A stemmed point is a projectile point made out of an elongated flake or blade with a slight retouch on the proximal end to the shape of an acute tip, and on the distal end to make a stem, which connects to a shaft. Stemmed points were the first composite tool types to appear on the Korean Peninsula, and were a symbol of new hunting strategies there and in adjacent regions [@seong2008tanged; @lee2019were]. A number of technological innovations are evident on stemmed points, including blade technology, multiple manufacturing stages, and evidence of resharpening and reusing, which were rarely seen earlier [@chang2013human; @bamforth2009projectile; @seong2015diversity]. Currently, the oldest stemmed points in Northeast Asia are from the Yonghodoing site in Korea, dating back to 38.5ka [@seong2008tanged; @seong2015diversity; @bae2012nature]. Following their appearance in Korea, stemmed points spread to the Japanese archipelago [@chang2013human].

Previous studies of the Late Paleolithic technological transitions in Korea have mainly focused on the possible origin locations of stemmed points as part of the discussion of the timing and routes of modern human dispersal in eastern Asia. The debate about the origin of stemmed points can be summarized into two competing models: *in situ* evolution [@seong2009emergence] and heterogenic migration [@bae2010origin]. The *in situ* model claims that stemmed points and other Late Paleolithic assemblages, including blade industries, autonomously emerged in the Korean peninsula as a form of convergent evolution [@seong2006structure; @seong2008tanged; @seong2009emergence]. To support his claim, @seong2009emergence examined the blade-to-flake ratios of stone artifact assemblages in South Korea. In his view, the blade industry represents a new technology, while flakes indicate a continuously-used existing one. He argues that the increased ratio of blades in stone artifact assemblages during the Late Paleolithic shows an expansion of the new technology after its local invention. In addition, Seong claims that increased numbers of stemmed points over time, and standardization of their shape, supports the prominance of gradual, local, evolutionary processes in the emergence of new technologies.

By contrast, the migration model contends that the new blade industry includes stemmed points, and the earlier flake tool tradition, including large cores, polyhedrals, choppers, and handaxes, originated outside of Korea [@bae2010origin; @bae2012nature; @bae2013early]. While the *in situ* model claims that the heterogenic character is the result of indigenous development, the migration model proposes that the heterogeneity is the result of the continuous influx of modern human populations from both north and south. Specifically, the blade technology is claimed to have been introduced from Siberia, Mongolia, or other regions of northeast China following the Liaohe and Sunghe rivers around 35 ka BP, while the simple flake-tool tradition came from southern China [@bae2013early]. The migration of people from the southern route is supported by genetic studies of the Y chromosome, indicating that the O3-M122 M122 haplotype originated from southern East Asia and moved to northeastern Asia, including Korea, at 30-25 ka BP [@shi2005chromosome]. Bae et al. [-@bae2012nature] assume this southern migration could be related to paleoenvironmental fluctuations during the MIS 3 to 2 transition, which made the Yellow Sea/West Sea region open.

Lee [-@lee2013current] argues that the transition to the Korean Late Paleolithic might be more complicated than either models of migration or *in situ* development. He partly agrees with the *in situ* model that simple flake tools had continuously been used in Korea as the result of ancestor-descendant relationships under conditions of low effective population size. With regard to the blade industry, he claims that the low degrees of uniformity and small quantities of blade-associated toolkits indicate an origin outside of Korea, perhaps resulting from trade or migration. We explore these three options, *in situ*, migration, and a mixture of the two, by measuring transmission biases in assemblages of stemmed points.

@@

To measure transmission biases in tool-making, we draw on Darwinian evolutionary theory. This body of theory has helped archaeologists understand a variety of technological innovations and related human behaviors, such as using human behavioral ecology, phylogenetics, cultural transmission theory, and niche construction theory [@dunnell1980evolutionary; @richerson1992cultural; @lipo1997population; @bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural]. Cultural evolutionary approaches are effective at explaining cultural changes using mechanisms of inheritance, variation-generating processes, and selection. The mechanisms are similar to biological evolution, but in cultural systems they have unique and distinctive properties. Social transmission strongly affects these key evolutionary processes [@Whiten_2017] and can occur through various learning processes, such as stimulus enhancement, emulation, imitation, and teaching [@o2000applying; @schillinger2014copying]. We focus here on how different social contexts can result in different modes of social transmission of tool-making skills among hunter-gatherer populations.

Cultural transmission theory holds that information about social and technological behaviours is acquired through interaction with other individuals and the environment [@boyd1988culture; @richerson1992cultural]. Individuals learn by themselves (e.g. trial and error) or from each other by sharing information. Information can be modified (also known as 'biased') depending on an individual's transmission context and cultural repertoire. Modification of information can occur by recombination, loss, or partial alteration [@eerkens2005cultural; @o2017dual]. Transmission biases can be important loci of changes in material culture, and can be influenced by the social contexts of cultural transmission [@heyes1994social; @eerkens2007cultural; @lycett2015cultural; @creanza2017cultural; @kendal2018social]. These biases include guided variation (where individuals learn new behaviors through trial and error), content-based bias (where some aspect of the transmitted instructions, such as cultural preferences, makes them more likely to be adopted), frequency-based bias (where an individual is biased to choose particular behaviors based on their perceived frequency in the population, such as extremely popular or rare behaviors), and indirect bias (where a behavior is transmitted because of its association with other attributes, such as the prestige or skill of other individuals) [@boyd1988culture; @richerson1992cultural; @o2017dual].

@@

Following Bettinger and Eerkens' framework, Garvey [-@garvey2018current] uses simulation to explore the degree of standardization represented by coefficient of variation (CV) values of projectile points from the US Southwest and westernmost southern High Plains. Garvey measured CV values for weight, thickness, width and length on two types of projectile points, Washita and Fresno points, from the Henderson site. She then computed simulated CV values according to three scenarios of different levels of transmission fidelity (i.e. CV = 10%, 5%, and 3%). The observed metrics of archaeological projectile points are closest to simulated metrics with 3% CV, which represents "extremely high-fidelity copying." A CV of 3% is the Weber fraction, or threshold of human visual perception, with variation below this value being too subtle for people to notice [@eerkens2000practice]. Garvey's work demonstrates how the cultural transmission of tool-making behaviors can be measured, simulated, and interpreted from the archaeological record.

Garvey, Bettinger and Eerkens inferred transmission processes from material culture because they were unable to observe them directly. Direct observations have been reported by Mesoudi and O'Brien [-@mesoudi2008cultural], who conducted experimental research using groups of undergraduate students to test the underlying assumptions of Bettinger and Eerkens' study. They simulated model-based bias in projectile point-making by providing the design of a model and information about the model's prior success to their research participants. They also simulated guided variation by allowing their participants to explore their own designs. They observed that the majority of participants who were able to choose from the previous design ultimately copied the most-successful model. Metric attributes of projectile points made when copied from successful models were more highly correlated than attributes of points made by trial and error. Mesoudi and O'Brien's experimental results are critical because they confirm the robustness of previous assumptions about cultural transmission biases, and validate the connection between material culture variation and cultural transmission biases.

@@

Inspired by Bettinger and Eerkens' approach, we use two contrasting transmission modes, guided variation and indirect bias, to investigate the spread of stemmed point technology during the Korean Late Paleolithic period. We propose a spectrum on which we can locate foraging groups and how they started to make stemmed points, depending on their degree of social isolation or social connectedness. On one end of the spectrum we have socially isolated groups who made stemmed points through guided variation; and on the other end of the spectrum, we have socially connected groups whose knowledge of stemmed points derived from transmission processes dominated by indirect bias.

To contextualize this further, on one end of the spectrum we have socially isolated groups who stayed in physically remote places apart from other groups, or had unfamiliar neighbors with limited contact between groups. Our assumption for this social context is that individuals or groups acquired the technology of stemmed points by modifying existing flake tool forms through trial-and-error processes to solve problems relating to resource procurement (cf. Seong's *in situ* model). If this social context was prevalent, we predict that this trial and error behavior could have left a distinctive signature on the metric variables of the stemmed points. If trial and error was the dominant bias in the transmission of knowledge about how to make stemmed points, we expect that the morphological attributes on a stemmed point will be poorly correlated and show high variation within assemblages and between sites.

On the other end of the spectrum, we have socially connected groups that occupied places close to other groups or had regular contact with them. Relatively high degrees of social connectivity provide frequent opportunities for observing others and acquiring information. Individuals or groups learned the technology of stemmed points by copying a model from another individual or group. In this scenario, it is likely that the model, or ideal, stemmed point has been highly successful and frequently chosen, and learners copy all information about the point design as a package. This implies a small number of locations, or a single point (perhaps from a source external to the Korean peninsula, cf. Bae's migration model), where stemmed points first appeared and then spread from the location(s). In this social context, indirect bias is the dominant influence on the transmission of lithic technology. If indirect bias was prominent in the Korean Paleolithic, we expect that stemmed points would be more standardized, attributes on the stemmed points would be more correlated, and assemblages from multiple sites would exhibit less variance between and within them.

@@

After the first discovery of stemmed points at the Seokjangri site in the 1960s, more than 450 have been found in nearly 30 sites across Korea (@fig-map) [@Sohn_1967; @lee2019were; @Chong_2021]. While most sites contain only a few points, a few sites have many more, such as Suyanggae (n = 55), Jingeuneul (n = 99), and Yongsandong (n = 38) [@Kim_2017]. Among these stemmed points, we selected those that were unbroken from the tip to the stem. We excluded artifacts that were recorded as stemmed points but lacked a stem. This resulted in a sample of `r total_number_of_sp` stemmed points from `r total_number_of_assemblages` assemblages unearthed from `r total_number_of_sites` sites spanning the period 40-17 ka (@tbl-sp-number). The images of the stemmed points were obtained through published excavation reports and direct photography during our research in local museums. We defined multiple assemblages from a site where artifact-bearing deposits were separated by culturally sterile deposits, or where distinct artifact-bearing stratigraphic units could be identified by major differences in the texture, color, and composition of the sedimentary deposits.

@@

To analyze morphological change over time, we used previously developed chronologies that identify three phases in the Korean Late Paleolithic [@seong2015diversity; @Park_2013]. These chronologies were constructed based on radiocarbon ages; assemblages without radiocarbon ages were classified by blades, stemmed point blanks, and toolkit composition. Following these previous chronological schemes, we divided the Korean Late Paleolithic assemblages into three chronological phases: 1) stemmed points made out of flakes and no blades in assemblages, 2) stemmed points made out of blades or flakes and the existence of blades in assemblages, and 3) stemmed points made out of blades and the existence of micro blades in assemblages (@tbl-korean-chronology). Applying this division is necessary because some sites such as Yonghodong, Goryeri, Jungmal, and Mungyeong have no radiocarbon ages due to research limitations. Inferring ages for these sites by analogy to the technological sequences at sites with dates is necessary for making maximal use of the available archaeological data to increase our sample size. We arranged our `r total_number_of_assemblages` assemblages containing stemmed points into these three phases to facilitate observations of change over time.

@@

Panel A of @fig-raw-materials-by-phase summarizes the amount of stemmed points in each chronological phase, and the distribution of raw materials across the phases. We exclude phase 1 from our analyses below because of the small sample size to compute correlations and CVs. Shale is the dominant raw material in both Phase 2 and 3. In Phase 3 there is an increase in the proportion of porphyry, driven largely by finds from Yongsangdong, and other raw materials. Panel B of @fig-raw-materials-by-phase shows the count of stemmed points by raw material type in assemblages with more than five stemmed points. Much of the raw material diversity comes from isolated finds, with those assemblages containing more than five points, showing high homogeneity in raw materials.

```{r}
#| label: fig-raw-materials-by-phase
#| fig-cap: "A: Three chronological phases of the Korean Late Paleolithic, and frequencies of stemmed points and distribution of raw materials in each phase. B: Frequencies of stemmed points by assemblage and raw material, for assemblages with five or more stemmed points."
#| fig-height: 4

@@

Following the metric attributes used in previous studies of cultural transmission and projectile points, i.e. maximum length and width [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we examined variations in shape and the relationship between each attribute using morphological attributes on stemmed points (@fig-labelled-schematic). We obtained our morphological attribute data from landmark analysis of digitized images of stemmed points. Compared to traditional caliper measurements, morphometric data yield interpretable numerical and visual outcomes more easily [@buchanan2010geometric; @thulman2012discriminating; @cardillo2016environment; @petvrik2018shape; @macleod2018quantitative; @okumura2019archaeology; @suarez2019life]. We did not include weight and thickness since these measurements were not available to us. For the landmark analysis, we placed 11 landmarks on the outline of each stemmed point and calculated distances between landmark coordinate pairs to derive attributes for statistical analysis. The landmarks we recorded are described in @tbl-landmark-abbreviations and shown in @fig-labelled-schematic. Using the point tool in ImageJ [@schneider2012nih], we captured the landmarks from images of the artifacts, and exported them as XY coordinate data for further analysis.

@@

We chose Coefficients of Variation (CV) to measure the variation among the measurements of the stemmed points. The CV is determined by the ratio of the standard deviation to the mean (usually expressed as a percentage). The method has been used in various disciplines to calculate standardization, precision, equality, homogeneity, etc. [@ng2006performance; @panichkitkosolkul2009improved; @panichkitkosolkul2013confidence; @wang2020standardization]. In archaeological studies, CV has been applied to measure the variation between artifacts and to test hypotheses about cultural evolutionary processes, including distinguishing between the types of learning biases affecting cultural transmission [@eerkens2001techniques; @eerkens2005cultural; @schillinger2014copying; @garvey2018current]. As a guide to interpreting CV values, Eerkens and Bettinger [-@eerkens2001techniques] claim that over 57.7% of the CV is the result of random production and below 1.7% is the byproduct of using a scale or template. Among all the various artifacts for which \@eerkens2001techniques summarized CV values, the most comparable case to stemmed points was Great Basin projectile points, whose CV values range from 6-55%, with an average of 22%. Similarly, projectile points from the US Southwest have CV values ranging from 11% to 33% [@garvey2018current].
